---
title: "Part 1: Differential Expression Analysis Across Drosophila Species"
author: "Candice Charles"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    number_sections: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
# Show code in the report and hide package startup noise
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This first report:

1. Imports and merges the six species count files.
2. Fits the DESeq2 model with `Dsim` as the reference.
3. Runs every species-versus-`Dsim` contrast.
4. Produces volcano plots and complete significant-gene heatmaps.
5. Saves the objects required by the separate GO enrichment report.

---

# Load Required Libraries

```{r load_libraries}
# Install CRAN packages if missing
cran_packages <- c("readxl", "dplyr", "ggplot2", "pheatmap", "tibble", "readr")
for (pkg in cran_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}

# Install Bioconductor packages if missing
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

bioc_packages <- c("DESeq2")
for (pkg in bioc_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
}

# Load libraries used in the analysis
library(DESeq2)
library(readxl)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(readr)

# Write a CSV normally, but use a timestamped filename if Windows locks the target.
safe_write_csv <- function(x, file, row.names = FALSE) {
  # Create the destination folder before attempting to write.
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)

  # Try the requested path and fall back safely when Excel or OneDrive locks it.
  tryCatch(
    {
      utils::write.csv(x, file, row.names = row.names)
      invisible(file)
    },
    error = function(e) {
      # Build an alternative filename that cannot overwrite an existing result.
      fallback <- file.path(
        dirname(file),
        paste0(
          tools::file_path_sans_ext(basename(file)),
          "_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".csv"
        )
      )

      # Save to the alternative path and report why it was used.
      utils::write.csv(x, fallback, row.names = row.names)
      warning("Original CSV was unavailable; saved instead as: ", fallback)
      invisible(fallback)
    }
  )
}
```

---

# Define File Paths

```{r file_paths}
# Update these paths if your input files move
files <- c(
  Dere = "C:/Users/Sophia C Charles/OneDrive - Queen Mary, University of London/Documents/Bioinformatics/Dere_counts.xlsx",
  Dmel = "C:/Users/Sophia C Charles/OneDrive - Queen Mary, University of London/Documents/Bioinformatics/Dmel_counts.xlsx",
  Dsuz = "C:/Users/Sophia C Charles/OneDrive - Queen Mary, University of London/Documents/Bioinformatics/Dsuz_counts.xlsx",
  Dsim = "C:/Users/Sophia C Charles/OneDrive - Queen Mary, University of London/Documents/Bioinformatics/Dsim_counts.xlsx",
  Dsec = "C:/Users/Sophia C Charles/OneDrive - Queen Mary, University of London/Documents/Bioinformatics/Dsec_counts.xlsx",
  Dsan = "C:/Users/Sophia C Charles/OneDrive - Queen Mary, University of London/Documents/Bioinformatics/Dsan_counts.xlsx"
)
```

---

# Data Import Function

```{r read_function}
# Read one species file, keep only gene IDs and count columns, and clean the table
read_species_counts <- function(file) {
  df <- read_excel(file, sheet = 1)

  count_cols <- setdiff(
    names(df),
    c("DmelID", grep("_gene_name$", names(df), value = TRUE))
  )

  df %>%
    dplyr::select(DmelID, dplyr::all_of(count_cols)) %>%
    dplyr::filter(!is.na(DmelID), DmelID != "") %>%
    dplyr::distinct(DmelID, .keep_all = TRUE) %>%
    dplyr::mutate(dplyr::across(dplyr::all_of(count_cols), as.numeric))
}
```

---

# Import and Merge All Species Data

```{r import_data}
# Read all count tables into a named list
count_list <- lapply(files, read_species_counts)

# Record which sample column belongs to which species
column_to_species <- list()
for (i in seq_along(count_list)) {
  species_name <- names(files)[i]
  sample_cols <- setdiff(colnames(count_list[[i]]), "DmelID")
  for (col in sample_cols) column_to_species[[col]] <- species_name
}

# Merge all species by DmelID and replace missing counts with 0
merged_counts <- Reduce(function(x, y) full_join(x, y, by = "DmelID"), count_list)
merged_counts[is.na(merged_counts)] <- 0

# Convert to a gene x sample count matrix
count_matrix <- as.data.frame(merged_counts)
rownames(count_matrix) <- count_matrix$DmelID
count_matrix$DmelID <- NULL
count_matrix[] <- lapply(count_matrix, function(x) as.integer(round(x)))

cat("Count matrix:", nrow(count_matrix), "genes x", ncol(count_matrix), "samples\n")
head(count_matrix[, 1:min(6, ncol(count_matrix))])
```

---

# Create Sample Metadata

```{r metadata}
# Build sample metadata using the original column-to-species mapping
sample_info <- data.frame(
  sample = colnames(count_matrix),
  species = sapply(colnames(count_matrix), function(x) column_to_species[[x]]),
  row.names = colnames(count_matrix)
)

# Set Dsim as the reference so every log2FC is relative to Dsim
sample_info$species <- factor(sample_info$species)
sample_info$species <- relevel(sample_info$species, ref = "Dsim")

# Make sure sample order matches the count matrix columns
stopifnot(all(rownames(sample_info) == colnames(count_matrix)))

print(sample_info)
print(table(sample_info$species))
```

---

# Create DESeq2 Dataset

```{r create_deseq}
# Build the DESeq2 object and remove genes with almost no information
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = sample_info,
  design = ~ species
)

dds <- dds[rowSums(counts(dds)) >= 10, ]
cat("Genes retained after filtering:", nrow(dds), "\n")
```

---

# Run DESeq2 Analysis

```{r run_deseq}
# Estimate size factors, dispersions, and fit the DE model
dds <- DESeq(dds)

# Save normalized counts for later inspection
norm_counts <- counts(dds, normalized = TRUE)
safe_write_csv(
  as.data.frame(norm_counts),
  "normalized_counts_all_species.csv",
  row.names = TRUE
)

# Use variance-stabilised counts for the heatmaps.
vsd_mat <- assay(vst(dds, blind = FALSE))

print(resultsNames(dds))
```

---

# Run All Pairwise Comparisons Against Dsim

```{r pairwise_results}
# Compare every non-Dsim species against Dsim
comparison_species <- setdiff(levels(sample_info$species), "Dsim")

# Use slightly relaxed cut-offs so downstream overlap and GO analyses have more genes
padj_cutoff <- 0.10
lfc_cutoff <- 0.50

# Store one result table per contrast
results_list <- list()
summary_table <- list()

for (sp in comparison_species) {
  # Positive log2FC means higher in the other species than in Dsim
  res <- results(dds, contrast = c("species", sp, "Dsim"))
  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange)))

  # Add simple flags that make later intersection steps easier
  res_df <- res_df %>%
    dplyr::mutate(
      significant = !is.na(padj) & padj < padj_cutoff & abs(log2FoldChange) >= lfc_cutoff,
      higher_in_Dsim = significant & log2FoldChange <= -lfc_cutoff,
      lower_in_Dsim = significant & log2FoldChange >= lfc_cutoff
    )

  results_list[[sp]] <- res_df
  summary_table[[sp]] <- data.frame(
    comparison = paste0(sp, "_vs_Dsim"),
    significant_genes = sum(res_df$significant, na.rm = TRUE),
    higher_in_Dsim = sum(res_df$higher_in_Dsim, na.rm = TRUE),
    lower_in_Dsim = sum(res_df$lower_in_Dsim, na.rm = TRUE)
  )

  # Export each contrast table so results are easy to inspect outside R.
  safe_write_csv(
    res_df,
    paste0("DESeq2_", sp, "_vs_Dsim.csv"),
    row.names = FALSE
  )
}

# Join per-contrast summaries into one table
contrast_summary <- dplyr::bind_rows(summary_table)
contrast_summary
```

---

# Volcano Plots for All Comparisons

```{r volcano_plots, fig.width=8, fig.height=6}
# Draw one volcano plot using the same thresholds and direction flags as the analysis.
plot_volcano <- function(res_df, species, padj_cutoff, lfc_cutoff) {
  # Create a comparison-specific label for genes higher in the other species.
  higher_label <- paste("Higher in", species)

  # Calculate plotting significance and assign each gene to an expression category.
  plot_df <- res_df %>%
    dplyr::mutate(
      neg_log10_padj = ifelse(is.na(padj), 0, -log10(pmax(padj, 1e-300))),
      status = dplyr::case_when(
        higher_in_Dsim ~ "Higher in Dsim",
        lower_in_Dsim ~ higher_label,
        TRUE ~ "Not significant"
      )
    )

  # Limit the view to the central 99.8% without removing observations from analysis.
  x_limit <- max(
    lfc_cutoff * 1.5,
    stats::quantile(abs(plot_df$log2FoldChange), 0.998, na.rm = TRUE)
  )

  # Apply the same display rule to very large adjusted-p-value signals.
  y_limit <- max(
    -log10(padj_cutoff) * 1.5,
    stats::quantile(plot_df$neg_log10_padj, 0.998, na.rm = TRUE)
  )

  # Construct the volcano plot with D. simulans direction shown consistently.
  ggplot(plot_df, aes(x = log2FoldChange, y = neg_log10_padj, colour = status)) +
    geom_point(alpha = 0.45, size = 0.7) +
    geom_vline(
      xintercept = c(-lfc_cutoff, lfc_cutoff),
      linetype = "dashed",
      colour = "grey40"
    ) +
    geom_hline(
      yintercept = -log10(padj_cutoff),
      linetype = "dashed",
      colour = "grey40"
    ) +
    scale_colour_manual(
      values = stats::setNames(
        c("#D95F0E", "#1B9E77", "grey75"),
        c("Higher in Dsim", higher_label, "Not significant")
      )
    ) +
    coord_cartesian(xlim = c(-x_limit, x_limit), ylim = c(0, y_limit)) +
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(
      title = paste(species, "vs Dsim"),
      subtitle = paste0("padj < ", padj_cutoff, ", |log2FC| \u2265 ", lfc_cutoff),
      x = expression(log[2]~fold~change),
      y = expression(-log[10](adjusted~p-value)),
      colour = NULL
    )
}

# Print one volcano plot for every species-versus-D. simulans contrast.
for (sp in comparison_species) {
  print(plot_volcano(results_list[[sp]], sp, padj_cutoff, lfc_cutoff))
}
```

---

# Heatmaps for All Comparisons

```{r all_heatmaps, fig.width=9, fig.height=10}
# Create a folder for complete scaled heatmap matrices and gene-order tables.
heatmap_output_dir <- "all_significant_gene_heatmaps"
dir.create(heatmap_output_dir, recursive = TRUE, showWarnings = FALSE)

# Draw one heatmap per contrast using every significant gene from that contrast.
for (sp in comparison_species) {
  # Retrieve the complete differential-expression table for this comparison.
  res_df <- results_list[[sp]]

  # Keep all significant genes, ordered by adjusted p-value before clustering.
  heatmap_genes <- res_df %>%
    dplyr::filter(significant) %>%
    dplyr::arrange(padj) %>%
    dplyr::pull(gene)

  # Skip the plot cleanly if a contrast has fewer than two significant genes.
  if (length(heatmap_genes) < 2) {
    cat("Skipping heatmap for", sp, "because fewer than 2 significant genes were found.\n")
    next
  }

  # Keep samples from D. simulans and the focal comparison species.
  keep_samples <- rownames(sample_info)[sample_info$species %in% c("Dsim", sp)]
  heat_mat <- vsd_mat[heatmap_genes, keep_samples, drop = FALSE]

  # Remove zero-variance genes because row scaling would convert them to NA values.
  row_sd <- apply(heat_mat, 1, stats::sd, na.rm = TRUE)
  variable_rows <- is.finite(row_sd) & row_sd > 0
  heat_mat <- heat_mat[variable_rows, , drop = FALSE]

  # Skip safely if fewer than two variable genes remain after quality control.
  if (nrow(heat_mat) < 2) {
    cat("Skipping heatmap for", sp, "because fewer than 2 variable genes remain.\n")
    next
  }

  # Standardise each gene so heatmap colours represent relative expression.
  heat_mat <- t(scale(t(heat_mat)))

  # Export the complete scaled matrix so every plotted gene remains reviewable.
  safe_write_csv(
    as.data.frame(heat_mat),
    file.path(heatmap_output_dir, paste0(sp, "_vs_Dsim_ALL_significant_heatmap_matrix.csv")),
    row.names = TRUE
  )

  # Scale row labels automatically between 1 and 10 points for improved readability.
  heatmap_fontsize <- max(1, min(10, 600 / nrow(heat_mat)))

  # Draw the complete heatmap; clustering determines the displayed gene order.
  pheatmap(
    heat_mat,
    annotation_col = sample_info[keep_samples, "species", drop = FALSE],
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = heatmap_fontsize,
    border_color = NA,
    main = paste("All significant DE genes:", sp, "vs Dsim")
  )
}
```

---

# Save Inputs for Part 2

The second report loads one compact RDS file, so the DESeq2 model does not need
to be fitted again before GO enrichment.

```{r save_go_handoff}
# Store only the objects required by the separate GO enrichment report.
go_input_objects <- list(
  results_list = results_list,
  comparison_species = comparison_species,
  tested_genes = rownames(dds),
  padj_cutoff = padj_cutoff,
  lfc_cutoff = lfc_cutoff
)

# Save the reusable handoff file beside both R Markdown documents.
saveRDS(go_input_objects, "deseq2_GO_input_objects.rds")

# Confirm where the handoff file was saved and how many contrasts it contains.
data.frame(
  handoff_file = normalizePath(
    "deseq2_GO_input_objects.rds",
    winslash = "/",
    mustWork = FALSE
  ),
  tested_genes = length(go_input_objects$tested_genes),
  contrasts = length(go_input_objects$results_list)
)
```
