---
title: "02. GO Enrichment Across Drosophila Species"
author: "Candice Charles"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

# Introduction

This report loads the DESeq2 results produced by `01_deseq2_pairwise_SIM.Rmd`, creates differentially expressed, upregulated-in-D. simulans, and downregulated-in-D. simulans gene sets supported in a majority of contrasts, performs GO Biological Process enrichment, exports every result, and visualises all significant terms as size-ranked bar charts.

# Load Required Libraries

```{r}
required_packages <- c(
  "clusterProfiler", "enrichplot", "org.Dm.eg.db", "AnnotationDbi",
  "dplyr", "ggplot2", "stringr", "tibble"
)
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages) > 0) {
  stop(
    "Missing packages: ", paste(missing_packages, collapse = ", "),
    ". Restore the project environment with renv::restore()."
  )
}

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(stringr)
  library(clusterProfiler)
  library(org.Dm.eg.db)
  library(AnnotationDbi)
})

# Write a CSV normally, but use a timestamped filename if Windows locks the target.
safe_write_csv <- function(x, file, row.names = FALSE) {
  # Create the destination folder before attempting to write.
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)

  # Try the requested path and fall back safely when Excel or OneDrive locks it.
  tryCatch(
    {
      utils::write.csv(x, file, row.names = row.names)
      invisible(file)
    },
    error = function(e) {
      # Build a unique alternative filename for a locked output.
      fallback <- file.path(
        dirname(file),
        paste0(
          tools::file_path_sans_ext(basename(file)),
          "_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".csv"
        )
      )

      # Save the result to the fallback path and report the change.
      utils::write.csv(x, fallback, row.names = row.names)
      warning("Original CSV was unavailable; saved instead as: ", fallback)
      invisible(fallback)
    }
  )
}
```

# Load DESeq2 Results from script 01

```{r}
# Specify the handoff file created at the end of script 01.
go_input_file <- "deseq2_GO_input_objects.rds"

# Stop with a clear instruction if script 01 has not created the handoff file.
if (!file.exists(go_input_file)) {
  stop("Run 01_deseq2_pairwise.Rmd first; deseq2_GO_input_objects.rds was not found.")
}

# Load the saved DESeq2 results and tested-gene universe.
go_input_objects <- readRDS(go_input_file)
results_list <- go_input_objects$results_list
comparison_species <- go_input_objects$comparison_species
tested_genes_by_contrast <- go_input_objects$tested_genes_by_contrast

# Use one shared background containing only genes that were testable in every
# contrast. This conservative intersection is defensible for the three groups,
# which are defined across the same set of pairwise comparisons.
if (is.null(tested_genes_by_contrast) || !is.list(tested_genes_by_contrast)) {
  stop("The DESeq2 handoff lacks tested_genes_by_contrast; rerun report 01 first.")
}
tested_genes <- Reduce(intersect, tested_genes_by_contrast)
padj_cutoff <- go_input_objects$padj_cutoff
lfc_cutoff <- go_input_objects$lfc_cutoff

# Verify that the handoff contains the objects required for GO enrichment.
stopifnot(
  is.list(results_list),
  length(results_list) > 0,
  length(tested_genes_by_contrast) == length(results_list),
  length(tested_genes) > 0
)

# Display a compact handoff summary before beginning enrichment.
data.frame(
  tested_genes = length(tested_genes),
  tested_genes_by_contrast = paste(lengths(tested_genes_by_contrast), collapse = ", "),
  contrasts = length(results_list),
  adjusted_p_cutoff = padj_cutoff,
  log2_fold_change_cutoff = lfc_cutoff
)
```

# Define the Three Majority-Supported Cross-Contrast Gene Groups

```{r}
# Store all GO-related tables and figures in one folder to keep the project tidy.
go_output_dir <- "GO_three_groups_results"
dir.create(go_output_dir, recursive = TRUE, showWarnings = FALSE)

# Require support in at least half the contrasts, with a minimum of two contrasts.
# With five contrasts this is a 3-of-5 majority-support rule, not an intersection.
min_overlap <- min(length(results_list), max(2, ceiling(length(results_list) / 2)))

# Define the three groups once so the same reusable code can process each group.
genes_by_group <- list(
  majority_supported_DE = lapply(results_list, function(x) unique(x$gene[x$significant %in% TRUE])),
  upregulated_in_Dsim = lapply(results_list, function(x) unique(x$gene[x$higher_in_Dsim %in% TRUE])),
  downregulated_in_Dsim = lapply(results_list, function(x) unique(x$gene[x$lower_in_Dsim %in% TRUE]))
)

# Preserve contrast names inside every group for transparent intersection tracking.
genes_by_group <- lapply(
  genes_by_group,
  function(x) stats::setNames(x, names(results_list))
)

# Count the number of supporting contrasts for every gene in one group.
count_gene_support <- function(contrast_lists) {
  sort(table(unlist(contrast_lists, use.names = FALSE)), decreasing = TRUE)
}

# Apply the support-count function to all three groups.
support_counts <- lapply(genes_by_group, count_gene_support)

# Retain genes meeting the majority-support threshold for GO enrichment.
go_gene_sets <- lapply(
  support_counts,
  function(x) names(x)[as.integer(x) >= min_overlap]
)

# Also calculate strict intersections supported by every contrast for review.
strict_gene_sets <- lapply(genes_by_group, function(x) Reduce(intersect, x))

# Create a complete audit table showing exactly which contrasts support each gene.
make_support_table <- function(contrast_lists, group_name) {
  # Collect every unique gene observed in this group.
  all_genes <- sort(unique(unlist(contrast_lists, use.names = FALSE)))

  # Record the supporting contrast names for each gene.
  supporting <- vapply(
    all_genes,
    function(gene_id) paste(
      names(contrast_lists)[vapply(
        contrast_lists,
        function(x) gene_id %in% x,
        logical(1)
      )],
      collapse = "; "
    ),
    character(1)
  )

  # Return one reviewable row per gene.
  data.frame(
    Group = group_name,
    gene = all_genes,
    n_contrasts = as.integer(support_counts[[group_name]][all_genes]),
    supporting_contrasts = supporting,
    selected_for_GO = all_genes %in% go_gene_sets[[group_name]],
    strict_intersection = all_genes %in% strict_gene_sets[[group_name]],
    stringsAsFactors = FALSE
  )
}

# Combine the intersection audit tables for shared, upregulated, and downregulated genes.
support_table <- dplyr::bind_rows(lapply(
  names(genes_by_group),
  function(group_name) make_support_table(genes_by_group[[group_name]], group_name)
)) %>%
  dplyr::arrange(Group, dplyr::desc(n_contrasts), gene)

# Export the full gene-level intersection audit table.
safe_write_csv(
  support_table,
  file.path(go_output_dir, "three_groups_all_gene_support.csv"),
  row.names = FALSE
)

# Summarise relaxed and strict intersection sizes for immediate checking.
support_summary <- data.frame(
  Group = names(go_gene_sets),
  selected_genes = lengths(go_gene_sets),
  strict_all_contrast_genes = lengths(strict_gene_sets),
  minimum_required_contrasts = min_overlap
)

# Display and export the three-group intersection summary.
knitr::kable(support_summary, caption = "Majority-supported cross-contrast gene groups used for GO")
```

```{r}
safe_write_csv(
  support_summary,
  file.path(go_output_dir, "three_groups_support_summary.csv"),
  row.names = FALSE
)
```

# Prepare Gene IDs for GO Enrichment

```{r}
# Build one annotation row for every gene tested by DESeq2.
gene_table <- data.frame(gene = tested_genes, stringsAsFactors = FALSE) %>%
  dplyr::mutate(
    raw_id = gene,
    symbol_guess = sub("_OG.*$", "", gene)
  )

# Cache valid keys once so inapplicable candidate identifiers can be separated
# from genuine annotation failures before bitr() is called.
annotation_keys <- list(
  SYMBOL = AnnotationDbi::keys(org.Dm.eg.db, keytype = "SYMBOL"),
  FLYBASE = AnnotationDbi::keys(org.Dm.eg.db, keytype = "FLYBASE")
)

# Map only identifiers that are valid for the requested key type. An empty set of
# valid candidates is an ordinary no-mapping result; database failures still stop.
safe_bitr <- function(ids, id_type) {
  if (!id_type %in% names(annotation_keys)) {
    stop("Unsupported annotation key type: ", id_type)
  }

  ids <- unique(trimws(as.character(stats::na.omit(ids))))
  valid_ids <- intersect(ids[nzchar(ids)], annotation_keys[[id_type]])
  if (length(valid_ids) == 0) {
    empty <- data.frame(character(), character(), stringsAsFactors = FALSE)
    names(empty) <- c(id_type, "ENTREZID")
    return(empty)
  }

  tryCatch(
    suppressMessages(clusterProfiler::bitr(
      valid_ids,
      fromType = id_type,
      toType = "ENTREZID",
      OrgDb = org.Dm.eg.db
    )),
    error = function(e) {
      stop(
        "Gene-ID mapping failed for key type ", id_type,
        ": ", conditionMessage(e)
      )
    }
  )
}

# Map original genes to Entrez IDs using gene symbols first and FlyBase IDs second.
map_to_entrez <- function(gene_ids) {
  # Keep only annotation rows for the requested genes.
  requested <- gene_table %>%
    dplyr::filter(gene %in% unique(stats::na.omit(gene_ids)))

  # Return a stable empty table when the input gene set is empty.
  if (nrow(requested) == 0) {
    return(tibble::tibble(
      gene = character(),
      raw_id = character(),
      symbol_guess = character(),
      ENTREZID = character()
    ))
  }

  # Map the inferred symbol before the "_OG" suffix.
  symbol_map <- safe_bitr(requested$symbol_guess, "SYMBOL")

  # Map the complete identifier when it is already a FlyBase ID.
  flybase_map <- safe_bitr(requested$raw_id, "FLYBASE")

  # Join successful symbol mappings back to the original gene identifiers.
  symbol_hits <- if (nrow(symbol_map) > 0) {
    dplyr::inner_join(
      requested,
      symbol_map,
      by = c("symbol_guess" = "SYMBOL")
    )
  } else {
    tibble::tibble()
  }

  # Join successful FlyBase mappings back to the original gene identifiers.
  flybase_hits <- if (nrow(flybase_map) > 0) {
    dplyr::inner_join(
      requested,
      flybase_map,
      by = c("raw_id" = "FLYBASE")
    )
  } else {
    tibble::tibble()
  }

  # Combine both strategies and remove duplicated gene-to-Entrez mappings.
  dplyr::bind_rows(symbol_hits, flybase_hits) %>%
    dplyr::mutate(ENTREZID = as.character(ENTREZID)) %>%
    dplyr::distinct(gene, ENTREZID, .keep_all = TRUE)
}

# Map every tested gene to create the correct enrichment background.
universe_map <- map_to_entrez(tested_genes)

# Map all three cross-contrast gene groups with the same function.
group_maps <- lapply(go_gene_sets, map_to_entrez)

# Export unmapped IDs and one-to-many mappings for annotation review.
mapping_inputs <- c(list(background_universe = tested_genes), go_gene_sets)
mapping_outputs <- c(list(background_universe = universe_map), group_maps)
for (group_name in names(mapping_inputs)) {
  input_ids <- unique(stats::na.omit(mapping_inputs[[group_name]]))
  mapped <- mapping_outputs[[group_name]]
  unmapped <- data.frame(gene = setdiff(input_ids, unique(mapped$gene)))
  safe_write_csv(
    unmapped,
    file.path(go_output_dir, paste0("unmapped_", group_name, "_genes.csv"))
  )
  multiple_mappings <- mapped %>%
    dplyr::count(gene, name = "n_Entrez_IDs") %>%
    dplyr::filter(n_Entrez_IDs > 1) %>%
    dplyr::left_join(mapped, by = "gene")
  safe_write_csv(
    multiple_mappings,
    file.path(go_output_dir, paste0("multiple_Entrez_mappings_", group_name, ".csv"))
  )
}

# Summarise mapping performance because unmapped genes cannot contribute to GO.
mapping_diagnostics <- dplyr::bind_rows(
  data.frame(
    Group = "background_universe",
    input_genes = length(tested_genes),
    mapped_genes = dplyr::n_distinct(universe_map$gene)
  ),
  dplyr::bind_rows(lapply(names(go_gene_sets), function(group_name) {
    data.frame(
      Group = group_name,
      input_genes = length(go_gene_sets[[group_name]]),
      mapped_genes = dplyr::n_distinct(group_maps[[group_name]]$gene)
    )
  }))
) %>%
  dplyr::mutate(mapping_rate = round(mapped_genes / pmax(input_genes, 1), 3))

# Display and export the complete identifier-mapping diagnostics.
knitr::kable(mapping_diagnostics, caption = "GO identifier-mapping diagnostics")
```

```{r}
safe_write_csv(
  mapping_diagnostics,
  file.path(go_output_dir, "GO_mapping_diagnostics.csv"),
  row.names = FALSE
)

# Create one Entrez-to-symbol lookup for readable GO gene lists.
entrez_to_symbol <- AnnotationDbi::mapIds(
  org.Dm.eg.db,
  keys = unique(universe_map$ENTREZID),
  column = "SYMBOL",
  keytype = "ENTREZID",
  multiVals = "first"
)
```

```{r}
# Run GO Enrichment

go_fdr <- 0.05
min_genes_for_go <- 10

# Calculate GO semantic similarity once and reuse it for all three groups.
if (!requireNamespace("GOSemSim", quietly = TRUE)) {
  stop("Package GOSemSim is required for GO term simplification. Install it with BiocManager::install('GOSemSim').")
}
go_semdata <- GOSemSim::godata("org.Dm.eg.db", ont = "BP", computeIC = TRUE)

collapse_gene_symbols <- function(entrez_string) {
  entrez_ids <- strsplit(as.character(entrez_string), "/", fixed = TRUE)[[1]]
  symbols <- unname(entrez_to_symbol[entrez_ids])
  paste(unique(stats::na.omit(symbols)), collapse = "/")
}

# Annotate and optionally export a clusterProfiler result table.
annotate_go <- function(ego, group_name, write = FALSE, suffix = "") {
  tbl <- methods::slot(ego, "result")
  if (nrow(tbl) == 0) return(data.frame())
  tbl <- tbl %>%
    dplyr::mutate(
      Group = group_name,
      significant_BH = !is.na(p.adjust) & p.adjust <= go_fdr,
      geneSymbols = vapply(geneID, collapse_gene_symbols, character(1))
    ) %>%
    dplyr::arrange(dplyr::desc(Count), p.adjust, Description)
  if (write) {
    safe_write_csv(tbl, file.path(go_output_dir, paste0("GO_BP_", group_name, "_ALL_tested_terms", suffix, ".csv")), row.names = FALSE)
    safe_write_csv(dplyr::filter(tbl, significant_BH), file.path(go_output_dir, paste0("GO_BP_", group_name, "_ALL_significant_terms", suffix, ".csv")), row.names = FALSE)
  }
  tbl
}

run_go <- function(gene_map, group_name) {
  gene_ids <- unique(gene_map$ENTREZID)
  universe_ids <- unique(universe_map$ENTREZID)
  if (length(gene_ids) < min_genes_for_go || length(universe_ids) == 0) {
    cat("Skipping GO for", group_name, "- mapped genes:", length(gene_ids), "\n")
    return(data.frame())
  }
  ego <- suppressMessages(clusterProfiler::enrichGO(
    gene = gene_ids, universe = universe_ids, OrgDb = org.Dm.eg.db,
    keyType = "ENTREZID", ont = "BP", pAdjustMethod = "BH",
    pvalueCutoff = 1, qvalueCutoff = 1, readable = FALSE
  ))
  # Preserve all tested terms as an explicitly named raw audit trail.
  annotate_go(ego, group_name, write = TRUE, suffix = "_RAW")
  raw_tbl <- methods::slot(ego, "result")
  sig_rows <- !is.na(raw_tbl$p.adjust) & raw_tbl$p.adjust <= go_fdr
  if (!any(sig_rows)) return(data.frame())
  # Restrict similarity calculations to significant terms, following the documented order.
  ego_sig <- ego
  methods::slot(ego_sig, "result") <- raw_tbl[sig_rows, , drop = FALSE]
  ego_sig <- enrichplot::pairwise_termsim(ego_sig)
  ego_simple <- clusterProfiler::simplify(ego_sig, cutoff = 0.7, by = "p.adjust", select_fun = min, semData = go_semdata)
  # Export and return the significant, redundancy-reduced result used downstream.
  annotate_go(ego_simple, group_name, write = TRUE)
}
go_results <- lapply(names(group_maps), function(g) run_go(group_maps[[g]], g))
names(go_results) <- names(group_maps)

all_go_terms <- dplyr::bind_rows(go_results)
significant_go_terms <- dplyr::filter(all_go_terms, significant_BH)

safe_write_csv(all_go_terms, file.path(go_output_dir, "GO_BP_majority_supported_groups_ALL_tested_terms.csv"), row.names = FALSE)
safe_write_csv(significant_go_terms, file.path(go_output_dir, "GO_BP_majority_supported_groups_ALL_significant_terms.csv"), row.names = FALSE)

go_summary <- purrr::map_dfr(names(go_gene_sets), function(g) {
  res <- go_results[[g]]
  data.frame(
    Group = g,
    input_genes = length(go_gene_sets[[g]]),
    mapped_genes = dplyr::n_distinct(group_maps[[g]]$gene),
    tested_GO_terms = nrow(res),
    significant_GO_terms = if (nrow(res) > 0) sum(res$significant_BH) else 0
  )
})

knitr::kable(go_summary, caption = "GO enrichment summary for all three groups")
```

```{r}
safe_write_csv(
  go_summary,
  file.path(go_output_dir, "GO_BP_majority_supported_groups_summary.csv"),
  row.names = FALSE
)
```

# Preview Significant GO Terms

The table below previews up to 20 BH-significant terms per group. The CSV exports retain every significant and non-significant tested term for complete review.

```{r}
# Select the most useful review columns while retaining every significant row.
go_review_table <- if (nrow(significant_go_terms) > 0) {
  significant_go_terms %>%
    dplyr::group_by(Group) %>%
    dplyr::slice_head(n = 20) %>%
    dplyr::ungroup() %>%
    dplyr::select(
      Group,
      ID,
      Description,
      GeneRatio,
      BgRatio,
      Count,
      pvalue,
      p.adjust,
      qvalue,
      geneSymbols
    )
} else {
  data.frame(Note = "No GO terms passed the configured BH threshold.")
}

# Use a paginated table in HTML so long results remain searchable and readable.
if (knitr::is_html_output()) {
  rmarkdown::paged_table(go_review_table, options = list(rows.print = 20))
} else {
  knitr::kable(go_review_table, digits = 4)
}
```

# Export Every GO-Term-to-Gene Link and Show a Preview

```{r}
# Expand one significant GO result into one row per contributing original gene.
expand_go_genes <- function(go_table, mapped_genes, group_name) {
  # Return an empty table when this group has no significant terms.
  if (nrow(go_table) == 0 || !any(go_table$significant_BH)) return(data.frame())

  # Keep all significant terms and expand their slash-delimited Entrez lists.
  dplyr::bind_rows(lapply(
    which(go_table$significant_BH),
    function(i) {
      # Split the current GO term into individual Entrez identifiers.
      term_entrez <- strsplit(go_table$geneID[i], "/", fixed = TRUE)[[1]]

      # Link each Entrez ID back to every matching original gene identifier.
      mapped_genes %>%
        dplyr::filter(ENTREZID %in% term_entrez) %>%
        dplyr::distinct(gene, symbol_guess, ENTREZID) %>%
        dplyr::mutate(
          Group = group_name,
          GO_ID = go_table$ID[i],
          GO_Description = go_table$Description[i],
          GO_p_adjust = go_table$p.adjust[i],
          .before = 1
        )
    }
  ))
}

# Expand every significant term from all three groups with the same helper.
go_term_gene_details <- dplyr::bind_rows(lapply(
  names(go_results),
  function(group_name) expand_go_genes(
    go_results[[group_name]],
    group_maps[[group_name]],
    group_name
  )
))

# Sort the detailed table only when at least one term-to-gene link exists.
if (nrow(go_term_gene_details) > 0) {
  go_term_gene_details <- go_term_gene_details %>%
    dplyr::arrange(Group, GO_p_adjust, GO_ID, gene)
}

# Export the complete GO-term-to-gene audit table.
safe_write_csv(
  go_term_gene_details,
  file.path(go_output_dir, "GO_BP_ALL_significant_terms_with_ALL_genes.csv"),
  row.names = FALSE
)

# Display the detailed table in a paginated format when rows are available.
if (nrow(go_term_gene_details) > 0 && knitr::is_html_output()) {
  rmarkdown::paged_table(utils::head(go_term_gene_details, 100), options = list(rows.print = 20))
} else if (nrow(go_term_gene_details) > 0) {
  knitr::kable(go_term_gene_details, digits = 4)
} else {
  cat("No significant GO-term-to-gene links were available.\n")
}
```

# Visualise Representative Significant GO Terms

Each horizontal bar represents one of the 15 largest significant GO terms in a group. Bar length shows Count, the number of genes from the analysed group associated with that term. Complete significant-term and tested-term results remain available in the CSV exports.

```{r go-barplots-shared, fig.width=14, fig.height=8}
# Set pagination in one place so 10, 15, or 20 terms can be tried easily.
go_terms_per_page <- 15L

# Scale exported page height to its actual number of terms.
go_barplot_height <- function(number_of_terms) {
  max(4, 0.5 * number_of_terms + 2)
}

# Create paginated horizontal bar charts ordered by descending GO-term gene count.
plot_go_bars <- function(
  go_table,
  group_label,
  terms_per_page = go_terms_per_page
) {
  # Stop cleanly when enrichment did not return a result table.
  if (is.null(go_table) || nrow(go_table) == 0) {
    cat("No GO results for", group_label, "\n")
    return(invisible(NULL))
  }

  # Retain every significant term and rank terms from largest to smallest.
  significant_terms <- go_table %>%
    dplyr::filter(significant_BH) %>%
    dplyr::arrange(dplyr::desc(Count), p.adjust, Description) %>%
    dplyr::slice_head(n = 15)

  # Stop cleanly when no term passes the selected BH threshold.
  if (nrow(significant_terms) == 0) {
    cat("No significant GO terms for", group_label, "\n")
    return(invisible(NULL))
  }

  # Store the descending size rank used to paginate the complete term list.
  significant_terms$Size_rank <- seq_len(nrow(significant_terms))

  # Divide the ordered terms into readable pages without dropping any term.
  page_ids <- ceiling(significant_terms$Size_rank / terms_per_page)

  # Build one bar chart for each consecutive size-ranked page.
  for (page_id in unique(page_ids)) {
    # Select the next group of largest remaining GO terms.
    page_terms <- significant_terms[page_ids == page_id, , drop = FALSE]

    # Wrap descriptions for normal left-to-right reading on the vertical axis.
    page_terms$Term <- paste0(
      page_terms$ID,
      "\n",
      stringr::str_wrap(page_terms$Description, width = 42)
    )
    # Reverse levels so the largest term appears at the top after coord_flip().
    page_terms$Term <- factor(page_terms$Term, levels = rev(page_terms$Term))

    # Convert adjusted p-values to the existing positive colour scale.
    page_terms$MinusLog10FDR <- -log10(
      pmax(page_terms$p.adjust, .Machine$double.xmin)
    )

    # Draw horizontal bars: GO terms on the y-axis and gene count on the x-axis.
    bar_plot <- ggplot2::ggplot(
      page_terms,
      ggplot2::aes(x = Term, y = Count, fill = MinusLog10FDR)
    ) +
      ggplot2::geom_col(width = 0.78) +
      ggplot2::geom_text(
        ggplot2::aes(label = Count),
        hjust = -0.15,
        size = 4.5
      ) +
      ggplot2::scale_fill_gradient(low = "#BFD3E6", high = "#084594") +
      ggplot2::scale_y_continuous(
        expand = ggplot2::expansion(mult = c(0, 0.14))
      ) +
      ggplot2::coord_flip(clip = "off") +
      ggplot2::theme_minimal(base_size = 14) +
      ggplot2::theme(
        panel.grid.major.y = ggplot2::element_blank(),
        panel.grid.minor = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_text(size = 12),
        axis.text.y = ggplot2::element_text(size = 11),
        axis.title = ggplot2::element_text(size = 16),
        plot.title = ggplot2::element_text(size = 18, face = "bold"),
        plot.subtitle = ggplot2::element_text(size = 14),
        legend.title = ggplot2::element_text(size = 14),
        legend.text = ggplot2::element_text(size = 13),
        plot.margin = ggplot2::margin(8, 22, 8, 8)
      ) +
      ggplot2::labs(
        title = paste("GO Biological Process enrichment:", group_label),
        subtitle = paste0(
          "Page ", page_id, " of ", max(page_ids),
          "; size ranks ", min(page_terms$Size_rank),
          "\u2013", max(page_terms$Size_rank),
          " (largest to smallest by gene count)"
        ),
        x = "GO term",
        y = "Gene count",
        fill = expression(-log[10]("BH adjusted p-value"))
      )

    # Print each page once so it is embedded in the knitted HTML.
    print(bar_plot)

    # Give sparse final pages a shorter PNG while retaining a readable minimum.
    page_height <- go_barplot_height(nrow(page_terms))
    safe_group_label <- gsub("[^A-Za-z0-9]+", "_", group_label)
    safe_group_label <- gsub("^_+|_+$", "", safe_group_label)
    export_file <- file.path(
      go_output_dir,
      paste0("GO_barplot_", safe_group_label, "_page", page_id, ".png")
    )
    ggplot2::ggsave(
      filename = export_file,
      plot = bar_plot,
      width = 14,
      height = page_height,
      dpi = 300
    )
  }
}

# Plot all significant shared-DE GO terms as size-ranked bar charts.
plot_go_bars(
  go_results$majority_supported_DE,
  "genes supported in a majority of contrasts"
)
```

```{r go-barplots-downregulated, fig.width=14, fig.height=8}
# Plot all terms from genes repeatedly lower in D. simulans.
plot_go_bars(
  go_results$downregulated_in_Dsim,
  "downregulated in D. simulans"
)
```

```{r go-barplots-upregulated, fig.width=14, fig.height=8}
# Plot all terms from genes repeatedly higher in D. simulans.
plot_go_bars(
  go_results$upregulated_in_Dsim,
  "upregulated in D. simulans"
)
```

<details>
<summary>R session information</summary>

```{r session_info}
sessionInfo()
```

</details>








