---
title: "01. DESeq2 Pairwise Differential Expression - D. erecta"
author: "Candice Charles"
date: "2026-07-01"
params:
  count_dir: "D:/Bioinformatics"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    number_sections: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
  mathjax: null
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
# Show code in the report and hide package startup noise
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

1. Imports and merges the six species count files.
2. Fits the DESeq2 model with `Dere` as the reference.
3. Runs every species-versus-`Dere` contrast.
4. Produces volcano plots and complete significant-gene heatmaps.
5. Saves the objects required by the separate GO enrichment report.

---

The strict sensitivity result calls `results()` a second time because DESeq2 re-optimises independent filtering for each alpha; simply re-thresholding the original adjusted p-values would not be equivalent.

## Load Required Libraries

```{r load_libraries}
required_packages <- c("DESeq2", "apeglm", "readxl", "dplyr", "ggplot2", "pheatmap", "tibble", "readr")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) stop("Missing packages: ", paste(missing_packages, collapse = ", "), ". Restore the project environment with renv::restore().")
suppressPackageStartupMessages({ library(DESeq2); library(apeglm); library(readxl); library(dplyr); library(ggplot2); library(pheatmap); library(tibble); library(readr) })

# Write a CSV normally, but use a timestamped filename if Windows locks the target.
safe_write_csv <- function(x, file, row.names = FALSE) {
  # Create the destination folder before attempting to write.
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)

  # Try the requested path and fall back safely when Excel or OneDrive locks it.
  tryCatch(
    {
      utils::write.csv(x, file, row.names = row.names)
      invisible(file)
    },
    error = function(e) {
      # Build an alternative filename that cannot overwrite an existing result.
      fallback <- file.path(
        dirname(file),
        paste0(
          tools::file_path_sans_ext(basename(file)),
          "_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".csv"
        )
      )

      # Save to the alternative path and report why it was used.
      utils::write.csv(x, fallback, row.names = row.names)
      warning("Original CSV was unavailable; saved instead as: ", fallback)
      invisible(fallback)
    }
  )
}
```

---

## Define File Paths

```{r file_paths}
# Raw count workbooks are excluded from Git. They are currently stored in
# D:/Bioinformatics. Override `params$count_dir` when rendering if they move.
files <- c(
  Dere = file.path(params$count_dir, "Dere_counts.xlsx"),
  Dmel = file.path(params$count_dir, "Dmel_counts.xlsx"),
  Dsuz = file.path(params$count_dir, "Dsuz_counts.xlsx"),
  Dsim = file.path(params$count_dir, "Dsim_counts.xlsx"),
  Dsec = file.path(params$count_dir, "Dsec_counts.xlsx"),
  Dsan = file.path(params$count_dir, "Dsan_counts.xlsx")
)

# Stop before importing anything and report every missing workbook together.
missing_count_files <- files[!file.exists(files)]
if (length(missing_count_files) > 0) {
  stop(
    paste0(
      "Missing raw count workbook(s):
- ",
      paste(
        normalizePath(missing_count_files, winslash = "/", mustWork = FALSE),
        collapse = "
- "
      ),
      "
Set the YAML parameter `count_dir` to the directory containing all six files."
    ),
    call. = FALSE
  )
}
```

---

## Data Import Function

```{r read_function}
# Read one species file, keep only gene IDs and count columns, and clean the table
read_species_counts <- function(file) {
  df <- read_excel(file, sheet = 1)
  if (!("DmelID" %in% names(df))) stop("Count file lacks DmelID: ", file)
  count_cols <- setdiff(names(df), c("DmelID", grep("_gene_name$", names(df), value = TRUE)))
  if (length(count_cols) != 3L) stop("Expected exactly 3 replicate columns in ", file, "; found ", length(count_cols), ".")
  raw_values <- lapply(df[count_cols], function(v) { txt <- trimws(as.character(v)); txt[is.na(txt) | txt %in% c("", "NA", "NaN", "NULL")] <- NA_character_; txt })
  before_na <- sum(is.na(as.data.frame(raw_values)))
  converted <- lapply(raw_values, function(v) suppressWarnings(as.numeric(v)))
  after_na <- sum(is.na(as.data.frame(converted)))
  if (after_na > before_na) stop("Non-numeric values introduced during count conversion in ", file)
  out <- cbind(df["DmelID"], as.data.frame(converted)); names(out) <- c("DmelID", count_cols)
  out <- out[!is.na(out$DmelID) & out$DmelID != "", , drop=FALSE]
  has_data <- apply(out[count_cols], 1, function(x) any(!is.na(x) & x != 0))
  out <- out[has_data & !duplicated(out$DmelID), , drop = FALSE]
  out
}
```

---

## Import and Merge All Species Data

```{r import_data}
# Read all count tables into a named list
count_list <- lapply(files, read_species_counts)

# Record which sample column belongs to which species
column_to_species <- list()
for (i in seq_along(count_list)) {
  species_name <- names(files)[i]
  sample_cols <- setdiff(colnames(count_list[[i]]), "DmelID")
  for (col in sample_cols) column_to_species[[col]] <- species_name
}

# Merge species by DmelID and retain only complete cases.
merged_counts <- Reduce(function(x, y) full_join(x, y, by = "DmelID"), count_list)
missing_by_species <- data.frame(DmelID=merged_counts$DmelID, stringsAsFactors=FALSE)
for (sp in names(files)) { cols <- names(column_to_species)[vapply(column_to_species, identical, logical(1), sp)]; missing_by_species[[sp]] <- if(length(cols)) rowSums(is.na(merged_counts[cols])) > 0 else FALSE }
excluded_genes <- missing_by_species[rowSums(missing_by_species[-1]) > 0, , drop=FALSE]
safe_write_csv(excluded_genes, "genes_excluded_incomplete_cases_erecta.csv")

ortholog_loss_by_species <- data.frame(
  species = names(files),
  genes_missing = vapply(names(files), function(sp) sum(excluded_genes[[sp]]), integer(1)),
  percent_of_excluded = 100 * vapply(names(files), function(sp) sum(excluded_genes[[sp]]), integer(1)) / max(1, nrow(excluded_genes)),
  total_library_size = vapply(names(files), function(sp) {
    cols <- names(column_to_species)[vapply(column_to_species, identical, logical(1), sp)]
    sum(as.matrix(count_list[[sp]][, cols, drop=FALSE]), na.rm=TRUE)
  }, numeric(1)),
  stringsAsFactors = FALSE
)
knitr::kable(ortholog_loss_by_species, caption="Ortholog-loss diagnostic by species")
safe_write_csv(ortholog_loss_by_species, "ortholog_loss_by_species_erecta.csv")
merged_counts <- merged_counts[rowSums(is.na(merged_counts[-1])) == 0, , drop=FALSE]

cat("Excluded incomplete genes:", nrow(excluded_genes), "; retained complete genes:", nrow(merged_counts), "\
")

# Convert to a gene x sample count matrix
count_matrix <- as.data.frame(merged_counts)
rownames(count_matrix) <- count_matrix$DmelID
count_matrix$DmelID <- NULL
count_matrix[] <- lapply(count_matrix, function(x) as.integer(round(x)))

cat("Count matrix:", nrow(count_matrix), "genes x", ncol(count_matrix), "samples
")
```

The phylogenetic expectation is that D. santomea, as the closest relative of D. erecta in this panel, should show the fewest missing ortholog counts. The contrast-level DE counts instead place D. santomea third, while D. melanogaster has the fewest differences. The table above reports whether missing-gene counts and total library sizes show a species-specific pattern consistent with a reference-genome effect; it does not by itself establish the cause.

---

## Create Sample Metadata

```{r metadata}
# Build sample metadata using the original column-to-species mapping
sample_info <- data.frame(
  sample = colnames(count_matrix),
  species = sapply(colnames(count_matrix), function(x) column_to_species[[x]]),
  row.names = colnames(count_matrix)
)

# Set Dere as the reference so every log2FC is relative to Dere
sample_info$species <- factor(sample_info$species)
sample_info$species <- relevel(sample_info$species, ref = "Dere")

# Make sure sample order matches the count matrix columns
stopifnot(all(rownames(sample_info) == colnames(count_matrix)))

print(sample_info)
print(table(sample_info$species))
```

---

## Create DESeq2 Dataset

```{r create_deseq}
# Build the DESeq2 object and remove genes with almost no information
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = sample_info,
  design = ~ species
)

smallestGroupSize <- min(table(sample_info$species))
dds <- dds[rowSums(counts(dds) >= 10) >= smallestGroupSize, ]
cat("Group-aware filter retained:", nrow(dds), "genes")
```

---

```{r quality_control, fig.cap="Variance-stabilised PCA for the filtered count matrix.", fig.alt="PCA plot showing sample separation by species."}
vsd_qc <- vst(estimateSizeFactors(dds), blind = TRUE)
print(plotPCA(vsd_qc, intgroup = "species"))
```
## Run DESeq2 Analysis

```{r run_deseq}
# Estimate size factors, dispersions, and fit the DE model
dds <- DESeq(dds)

# Save normalized counts for later inspection
norm_counts <- counts(dds, normalized = TRUE)
safe_write_csv(
  as.data.frame(norm_counts),
  "normalized_counts_all_species2.csv",
  row.names = TRUE
)

# Use variance-stabilised counts for the heatmaps.
vsd_mat <- assay(vst(dds, blind = FALSE))

print(resultsNames(dds))
```

---

## Quality-control diagnostics

The DESeq2 results use a direct threshold test with `lfcThreshold`: genes are called significant only when the adjusted p-value is below the FDR cutoff and the data support an absolute log2 fold change greater than the specified effect-size threshold. The exported `log2FoldChange` column contains apeglm-shrunken estimates, while `raw_log2FoldChange` preserves the unshrunken MLE for auditability. Among significant genes, direction flags are assigned from the sign of the shrunken estimate; the threshold is not applied a second time.

## Run All Pairwise Comparisons Against Dere

```{r pairwise_results}
comparison_species <- setdiff(levels(sample_info$species), "Dere")
padj_cutoff <- 0.10
strict_padj_cutoff <- 0.05
strict_lfc_cutoff <- 1.00
lfc_cutoff <- 0.50
results_list <- list(); summary_table <- list(); tested_genes_by_contrast <- list()
for (sp in comparison_species) {
  coef_name <- grep(paste0("^species_", sp, "_vs_Dere$"), resultsNames(dds), value = TRUE)
  if (length(coef_name) != 1L) {
    stop("Expected exactly one coefficient matching ", sp, " vs the reference level; found ", length(coef_name), ". apeglm requires `coef`, so the focal species must be the reference level (check the relevel() call).")
  }
  raw_mle <- results(dds, name = coef_name, alpha = padj_cutoff)
  res <- results(dds, name = coef_name, alpha = padj_cutoff, lfcThreshold = lfc_cutoff, altHypothesis = "greaterAbs")
  strict_res <- results(dds, name = coef_name, alpha = strict_padj_cutoff, lfcThreshold = strict_lfc_cutoff, altHypothesis = "greaterAbs")
  shrunk <- lfcShrink(dds, coef = coef_name, res = raw_mle, type = "apeglm")
  res_df <- as.data.frame(res) %>% rownames_to_column("gene")
  raw_df <- as.data.frame(raw_mle) %>% rownames_to_column("gene") %>% dplyr::select(gene, raw_log2FoldChange = log2FoldChange, raw_pvalue = pvalue, raw_padj = padj)
  shr_df <- as.data.frame(shrunk) %>% rownames_to_column("gene") %>% dplyr::select(gene, log2FoldChange)
  strict_df <- as.data.frame(strict_res) %>% rownames_to_column("gene") %>% dplyr::select(gene, strict_padj = padj)
  res_df <- res_df %>% dplyr::rename(threshold_log2FoldChange = log2FoldChange, threshold_pvalue = pvalue, threshold_padj = padj) %>% dplyr::left_join(raw_df, by = "gene") %>% dplyr::left_join(shr_df, by = "gene") %>% dplyr::left_join(strict_df, by = "gene") %>% dplyr::mutate(pvalue = threshold_pvalue, padj = threshold_padj, significant = !is.na(threshold_padj) & threshold_padj < padj_cutoff, significant_strict = !is.na(strict_padj) & strict_padj < strict_padj_cutoff, higher_in_Dere = significant & log2FoldChange < 0, lower_in_Dere = significant & log2FoldChange > 0) %>% dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange)))
  results_list[[sp]] <- res_df
  tested_genes_by_contrast[[sp]] <- res_df$gene[!is.na(res_df$threshold_pvalue)]
  summary_table[[sp]] <- data.frame(comparison = paste0(sp, "_vs_Dere"), genes_tested = sum(!is.na(res_df$threshold_pvalue)), cook_distance_outliers = sum(is.na(raw_mle$pvalue) & !is.na(raw_mle$log2FoldChange)), low_count_independent_filtering = sum(is.na(raw_mle$pvalue) & is.na(raw_mle$log2FoldChange)), significant_genes = sum(res_df$significant, na.rm = TRUE), significant_genes_strict = sum(res_df$significant_strict, na.rm = TRUE), higher_in_Dere = sum(res_df$higher_in_Dere, na.rm = TRUE), lower_in_Dere = sum(res_df$lower_in_Dere, na.rm = TRUE))
  safe_write_csv(res_df, paste0("DESeq2_", sp, "_vs_Dere.csv"), row.names = FALSE)
}
contrast_summary <- dplyr::bind_rows(summary_table)
stopifnot(all(contrast_summary$significant_genes == contrast_summary$higher_in_Dere + contrast_summary$lower_in_Dere))
zero_shrunken_by_contrast <- dplyr::bind_rows(lapply(names(results_list), function(sp) data.frame(comparison = paste0(sp, "_vs_Dere"), significant_shrunken_log2FC_exactly_zero = sum(results_list[[sp]]$significant & results_list[[sp]]$log2FoldChange == 0, na.rm = TRUE))))
zero_shrunken_by_contrast
contrast_summary
```
```{r volcano_plots, fig.width=8, fig.height=6, fig.cap="Volcano plots for each species relative to D. erecta.", fig.alt="Volcano plots showing log2 fold changes and adjusted p-values for each species-versus-D. erecta contrast."}
# Full scientific names used in plot labels.
species_names <- c(
  Dere = "D. erecta",
  Dmel = "D. melanogaster",
  Dsim = "D. simulans",
  Dsec = "D. sechellia",
  Dsan = "D. santomea",
  Dsuz = "D. suzukii"
)

# Set these three values according to the script's reference species.
reference_code <- "Dere"
reference_label <- species_names[[reference_code]]
higher_reference_column <- "higher_in_Dere"
lower_reference_column <- "lower_in_Dere"

plot_volcano <- function(x, sp) {

  comparison_label <- species_names[[sp]]

  d <- x %>%
    mutate(
      y = -log10(pmax(padj, 1e-300)),
      status = case_when(
        .data[[higher_reference_column]] ~
          paste("Higher in", reference_label),
        .data[[lower_reference_column]] ~
          paste("Higher in", comparison_label),
        TRUE ~ "Not significant"
      )
    )

  plot_title <- bquote(
    italic(.(reference_label)) ~ "vs" ~ italic(.(comparison_label))
  )

  ggplot(d, aes(log2FoldChange, y, colour = status)) +
    geom_point(alpha = 0.80, size = 2.5) +
    geom_vline(
      xintercept = c(-lfc_cutoff, lfc_cutoff),
      linetype = "dashed",
      colour = "grey40"
    ) +
    geom_hline(
      yintercept = -log10(padj_cutoff),
      linetype = "dashed",
      colour = "red4"
    ) +
    theme_minimal(base_size = 16) +
    labs(
      title = plot_title,
      x = "Shrunken log2 fold change",
      y = "-log10 adjusted p-value",
      colour = NULL
    ) +
    theme(
      plot.title = element_text(size = 16),
      axis.title = element_text(size = 16),
      axis.text = element_text(size = 14),
      legend.text = element_text(size = 16)
    ) +
    guides(colour = guide_legend(override.aes = list(size = 3)))
}

for (sp in comparison_species) {
  print(plot_volcano(results_list[[sp]], sp))
}
```

---

## Heatmaps for All Comparisons

```{r all_heatmaps, fig.width=9, fig.height=10, fig.cap="Top-75 significant-gene heatmaps; complete matrices are supplied as CSV files for each D. erecta contrast.", fig.alt="Scaled expression heatmaps for significant genes in each species-versus-D. erecta comparison."}
# Create a folder for complete scaled heatmap matrices and gene-order tables.
heatmap_output_dir <- "all_significant_gene_heatmaps2"
dir.create(heatmap_output_dir, recursive = TRUE, showWarnings = FALSE)

# Draw one heatmap per contrast using every significant gene from that contrast.
for (sp in comparison_species) {
  # Retrieve the complete differential-expression table for this comparison.
  res_df <- results_list[[sp]]

  # Keep all significant genes, ordered by adjusted p-value before clustering.
  heatmap_genes <- res_df %>%
    dplyr::filter(significant) %>%
    dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange))) %>%
    dplyr::pull(gene)

  # Skip the plot cleanly if a contrast has fewer than two significant genes.
  if (length(heatmap_genes) < 2) {
    cat("Skipping heatmap for", sp, "because fewer than 2 significant genes were found.
")
    next
  }

  # Keep samples from D. erecta and the focal comparison species.
  keep_samples <- rownames(sample_info)[sample_info$species %in% c("Dere", sp)]
  heat_mat <- vsd_mat[heatmap_genes, keep_samples, drop = FALSE]

  # Remove zero-variance genes because row scaling would convert them to NA values.
  row_sd <- apply(heat_mat, 1, stats::sd, na.rm = TRUE)
  variable_rows <- is.finite(row_sd) & row_sd > 0
  heat_mat <- heat_mat[variable_rows, , drop = FALSE]

  # Skip safely if fewer than two variable genes remain after quality control.
  if (nrow(heat_mat) < 2) {
    cat("Skipping heatmap for", sp, "because fewer than 2 variable genes remain.
")
    next
  }

  # Standardise each gene so heatmap colours represent relative expression.
  heat_mat <- t(scale(t(heat_mat)))

  # Export the complete scaled matrix so every plotted gene remains reviewable.
  safe_write_csv(
    as.data.frame(heat_mat),
    file.path(heatmap_output_dir, paste0(sp, "_vs_Dere_ALL_significant_heatmap_matrix.csv")),
    row.names = TRUE
  )

  # The CSV above retains every significant gene; the plotted heatmap is capped at the top 75 for readability.
  heat_mat <- heat_mat[seq_len(min(75, nrow(heat_mat))), , drop = FALSE]
  # Scale row labels automatically between 1 and 10 points for improved readability.
  heatmap_fontsize <- max(1, min(10, 600 / nrow(heat_mat)))

  # Draw the complete heatmap; clustering determines the displayed gene order.
  heatmap_obj <- pheatmap(
    heat_mat,
    annotation_col = sample_info[keep_samples, "species", drop = FALSE],
    show_rownames = TRUE, show_colnames = TRUE, fontsize_row = heatmap_fontsize, border_color = NA,
    main = paste("Top 75 significant DE genes:", sp, "vs Dere"), silent = TRUE
  )
  grid::grid.newpage(); grid::grid.draw(heatmap_obj$gtable)
  png(file.path(heatmap_output_dir, paste0(sp, "_vs_Dere_ALL_significant_heatmap.png")), width=900, height=1000, res=120)
  grid::grid.newpage(); grid::grid.draw(heatmap_obj$gtable); invisible(dev.off())
}
```

---

## Save Inputs for Part 2 - Erecta

The second report loads one compact RDS file, so the DESeq2 model does not need
to be fitted again before GO enrichment.

```{r save_go_handoff}
# Store only the objects required by the separate GO enrichment report.
go_input_objects <- list(
  results_list = results_list,
  comparison_species = comparison_species,
  tested_genes_by_contrast = tested_genes_by_contrast,
  padj_cutoff = padj_cutoff,
  lfc_cutoff = lfc_cutoff
)

# Save the reusable handoff file beside both R Markdown documents.
saveRDS(go_input_objects, "deseq2_GO_input_objects.rds2")

# Confirm where the handoff file was saved and how many contrasts it contains.
data.frame(
  handoff_file = normalizePath(
    "deseq2_GO_input_objects.rds2",
    winslash = "/",
    mustWork = FALSE
  ),
  tested_genes_by_contrast = vapply(go_input_objects$tested_genes_by_contrast, length, integer(1)),
  contrasts = length(go_input_objects$results_list)
)
```


## Reproducibility

<details><summary>Session information</summary>

```{r session_info}
sessionInfo()
```

</details>
