---
title: "GO Enrichment - Erecta"
author: "Candice Charles"
date: "2026-07-01"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    number_sections: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
  mathjax: null
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
# Show code while suppressing package startup messages and warnings in the report.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This second report loads the DESeq2 results produced by Part 1, creates the
majority-supported, upregulated-in-*D. erecta*, and downregulated-in-*D. erecta* gene
intersections, performs GO Biological Process enrichment, exports every result,
and visualises all significant terms as size-ranked bar charts.

Differential-expression calls use BH-adjusted p < 0.10, whereas GO enrichment uses BH-adjusted p ≤ 0.05 (`go_fdr`); the stricter GO threshold limits false-positive functional annotations after testing many ontology terms.

---

## Load Required Libraries

```{r load_libraries}
required_packages <- c("dplyr", "ggplot2", "stringr", "tibble", "clusterProfiler", "enrichplot", "org.Dm.eg.db", "AnnotationDbi", "GOSemSim")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) stop("Missing packages: ", paste(missing_packages, collapse = ", "), ". Restore the project environment with renv::restore().")

# Load the packages used by the GO workflow and visualisations.
library(dplyr)
library(ggplot2)
library(stringr)
library(clusterProfiler)
library(org.Dm.eg.db)
library(AnnotationDbi)

# Write a CSV normally, but use a timestamped filename if Windows locks the target.
safe_write_csv <- function(x, file, row.names = FALSE) {
  # Create the destination folder before attempting to write.
  dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)

  # Try the requested path and fall back safely when Excel or OneDrive locks it.
  tryCatch(
    {
      utils::write.csv(x, file, row.names = row.names)
      invisible(file)
    },
    error = function(e) {
      # Build a unique alternative filename for a locked output.
      fallback <- file.path(
        dirname(file),
        paste0(
          tools::file_path_sans_ext(basename(file)),
          "_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".csv"
        )
      )

      # Save the result to the fallback path and report the change.
      utils::write.csv(x, fallback, row.names = row.names)
      warning("Original CSV was unavailable; saved instead as: ", fallback)
      invisible(fallback)
    }
  )
}
```

---

## Load DESeq2 Results from Part 1

```{r load_go_handoff}
# Specify the handoff file created at the end of Part 1.
go_input_file <- "deseq2_GO_input_objects.rds2"

# Stop with a clear instruction if Part 1 has not created the handoff file.
if (!file.exists(go_input_file)) {
  stop("Run Part 1 first; the file deseq2_GO_input_objects.rds2 was not found.")
}

# Load the saved DESeq2 results and tested-gene universe.
go_input_objects <- readRDS(go_input_file)
results_list <- go_input_objects$results_list
comparison_species <- go_input_objects$comparison_species
tested_genes_by_contrast <- go_input_objects$tested_genes_by_contrast
tested_genes <- Reduce(intersect, tested_genes_by_contrast)
padj_cutoff <- go_input_objects$padj_cutoff
lfc_cutoff <- go_input_objects$lfc_cutoff

# Verify that the handoff contains the objects required for GO enrichment.
stopifnot(
  is.list(results_list),
  length(results_list) > 0,
  length(tested_genes_by_contrast) > 0,
  all(lengths(tested_genes_by_contrast) > 0),
  length(tested_genes) > 0
)

# Display a compact handoff summary before beginning enrichment.
data.frame(
  tested_genes = length(tested_genes),
  tested_genes_by_contrast = vapply(tested_genes_by_contrast, length, integer(1)),
  contrasts = length(results_list),
  adjusted_p_cutoff = padj_cutoff,
  log2_fold_change_cutoff = lfc_cutoff
)
```

---

## Define the Three Cross-Contrast Gene Groups

```{r overlap_genes}
# Store all GO-related tables and figures in one folder to keep the project tidy.
go_output_dir <- "GO_three_groups_results2"
dir.create(go_output_dir, recursive = TRUE, showWarnings = FALSE)

# Require support in at least half the contrasts, with a minimum of two contrasts.
min_overlap <- min(length(results_list), max(2, ceiling(length(results_list) / 2)))

# Define the three groups once so the same reusable code can process each group.
genes_by_group <- list(
  majority_supported_DE = lapply(results_list, function(x) unique(x$gene[x$significant %in% TRUE])),
  upregulated_in_Dere = lapply(results_list, function(x) unique(x$gene[x$higher_in_Dere %in% TRUE])),
  downregulated_in_Dere = lapply(results_list, function(x) unique(x$gene[x$lower_in_Dere %in% TRUE]))
)

# Preserve contrast names inside every group for transparent intersection tracking.
genes_by_group <- lapply(
  genes_by_group,
  function(x) stats::setNames(x, names(results_list))
)

# Count the number of supporting contrasts for every gene in one group.
count_gene_support <- function(contrast_lists) {
  sort(table(unlist(contrast_lists, use.names = FALSE)), decreasing = TRUE)
}

# Apply the support-count function to all three groups.
support_counts <- lapply(genes_by_group, count_gene_support)

# Retain genes meeting the relaxed intersection threshold for GO enrichment.
go_gene_sets <- lapply(
  support_counts,
  function(x) names(x)[as.integer(x) >= min_overlap]
)

# Also calculate strict intersections supported by every contrast for review.
strict_gene_sets <- lapply(genes_by_group, function(x) Reduce(intersect, x))

# Create a complete audit table showing exactly which contrasts support each gene.
make_intersection_table <- function(contrast_lists, group_name) {
  # Collect every unique gene observed in this group.
  all_genes <- sort(unique(unlist(contrast_lists, use.names = FALSE)))

  # Record the supporting contrast names for each gene.
  supporting <- vapply(
    all_genes,
    function(gene_id) paste(
      names(contrast_lists)[vapply(
        contrast_lists,
        function(x) gene_id %in% x,
        logical(1)
      )],
      collapse = "; "
    ),
    character(1)
  )

  # Return one reviewable row per gene.
  data.frame(
    Group = group_name,
    gene = all_genes,
    n_contrasts = as.integer(support_counts[[group_name]][all_genes]),
    supporting_contrasts = supporting,
    selected_for_GO = all_genes %in% go_gene_sets[[group_name]],
    strict_intersection = all_genes %in% strict_gene_sets[[group_name]],
    primary_comparison_supported = vapply(strsplit(supporting, "; ", fixed=TRUE), function(x) all(c("Dmel", "Dsan") %in% x), logical(1)),
    stringsAsFactors = FALSE
  )
}

# Combine the intersection audit tables for majority-supported, upregulated, and downregulated genes.
intersection_table <- dplyr::bind_rows(lapply(
  names(genes_by_group),
  function(group_name) make_intersection_table(genes_by_group[[group_name]], group_name)
)) %>%
  dplyr::arrange(Group, dplyr::desc(n_contrasts), gene)

# Export the full gene-level intersection audit table.
safe_write_csv(
  intersection_table,
  file.path(go_output_dir, "three_groups_all_gene_intersections2.csv"),
  row.names = FALSE
)

# Summarise relaxed and strict intersection sizes for immediate checking.
intersection_summary <- data.frame(
  Group = names(go_gene_sets),
  selected_genes = lengths(go_gene_sets),
  strict_all_contrast_genes = lengths(strict_gene_sets),
  minimum_required_contrasts = min_overlap
)

# Display and export the three-group intersection summary.
knitr::kable(intersection_summary, caption = "Cross-contrast gene groups used for GO")
safe_write_csv(
  intersection_summary,
  file.path(go_output_dir, "three_groups_intersection_summary2.csv"),
  row.names = FALSE
)
```

---

## Prepare Gene IDs for GO Enrichment

```{r prepare_go_ids}
# Build one annotation row for every gene tested by DESeq2.
gene_table <- data.frame(gene = tested_genes, stringsAsFactors = FALSE) %>%
  dplyr::mutate(
    raw_id = gene,
    symbol_guess = sub("_OG.*$", "", gene)
  )

# Map an identifier vector safely and return an empty table if mapping fails.
safe_bitr <- function(ids, id_type) {
  tryCatch(
    suppressMessages(clusterProfiler::bitr(
      unique(ids),
      fromType = id_type,
      toType = "ENTREZID",
      OrgDb = org.Dm.eg.db
    )),
    error = function(e) data.frame()
  )
}

# Map original genes to Entrez IDs using gene symbols first and FlyBase IDs second.
map_to_entrez <- function(gene_ids) {
  # Keep only annotation rows for the requested genes.
  requested <- gene_table %>%
    dplyr::filter(gene %in% unique(stats::na.omit(gene_ids)))

  # Return a stable empty table when the input gene set is empty.
  if (nrow(requested) == 0) {
    return(tibble::tibble(
      gene = character(),
      raw_id = character(),
      symbol_guess = character(),
      ENTREZID = character()
    ))
  }

  # Map the inferred symbol before the "_OG" suffix.
  symbol_map <- safe_bitr(requested$symbol_guess, "SYMBOL")

  # Map the complete identifier when it is already a FlyBase ID.
  flybase_map <- safe_bitr(requested$raw_id, "FLYBASE")

  # Join successful symbol mappings back to the original gene identifiers.
  symbol_hits <- if (nrow(symbol_map) > 0) {
    dplyr::inner_join(
      requested,
      symbol_map,
      by = c("symbol_guess" = "SYMBOL")
    )
  } else {
    tibble::tibble()
  }

  # Join successful FlyBase mappings back to the original gene identifiers.
  flybase_hits <- if (nrow(flybase_map) > 0) {
    dplyr::inner_join(
      requested,
      flybase_map,
      by = c("raw_id" = "FLYBASE")
    )
  } else {
    tibble::tibble()
  }

  # Combine both strategies and remove duplicated gene-to-Entrez mappings.
  dplyr::bind_rows(symbol_hits, flybase_hits) %>%
    dplyr::mutate(ENTREZID = as.character(ENTREZID)) %>%
    dplyr::distinct(gene, ENTREZID, .keep_all = TRUE)
}

# Map every tested gene to create the correct enrichment background.
universe_map <- map_to_entrez(tested_genes)

# Map all three cross-contrast gene groups with the same function.
group_maps <- lapply(go_gene_sets, map_to_entrez)

# Summarise mapping performance because unmapped genes cannot contribute to GO.
mapping_diagnostics <- dplyr::bind_rows(
  data.frame(
    Group = "background_universe",
    input_genes = length(tested_genes),
    mapped_genes = dplyr::n_distinct(universe_map$gene)
  ),
  dplyr::bind_rows(lapply(names(go_gene_sets), function(group_name) {
    data.frame(
      Group = group_name,
      input_genes = length(go_gene_sets[[group_name]]),
      mapped_genes = dplyr::n_distinct(group_maps[[group_name]]$gene)
    )
  }))
) %>%
  dplyr::mutate(mapping_rate = round(mapped_genes / pmax(input_genes, 1), 3))

# Display and export the complete identifier-mapping diagnostics.
knitr::kable(mapping_diagnostics, caption = "GO identifier-mapping diagnostics")
safe_write_csv(
  mapping_diagnostics,
  file.path(go_output_dir, "GO_mapping_diagnostics2.csv"),
  row.names = FALSE
)

# Create one Entrez-to-symbol lookup for readable GO gene lists.
entrez_to_symbol <- AnnotationDbi::mapIds(
  org.Dm.eg.db,
  keys = unique(universe_map$ENTREZID),
  column = "SYMBOL",
  keytype = "ENTREZID",
  multiVals = "first"
)
```

---

## Run GO Enrichment

```{r go_enrichment}
# Set the BH-adjusted p-value threshold used to call a GO term significant.
go_fdr <- 0.05

# Require enough mapped genes to avoid unstable enrichment tests.
min_genes_for_go <- 10

# Convert a slash-delimited Entrez list into readable gene symbols.
collapse_gene_symbols <- function(entrez_string) {
  # Split the clusterProfiler gene list into individual Entrez IDs.
  entrez_ids <- strsplit(as.character(entrez_string), "/", fixed = TRUE)[[1]]

  # Retrieve symbols from the lookup created once in the mapping section.
  symbols <- unname(entrez_to_symbol[entrez_ids])

  # Remove missing values and return one slash-delimited symbol list.
  paste(unique(stats::na.omit(symbols)), collapse = "/")
}

# Build the Biological Process semantic-data object once and reuse it for every
# group instead of recomputing information content inside each simplify() call.
go_semdata <- GOSemSim::godata("org.Dm.eg.db", ont = "BP", computeIC = TRUE)

term_counts <- list()
# Run the same enrichment, annotation, and export steps for any mapped gene group.
run_go <- function(gene_map, group_name) {
  # Retain unique group and background identifiers.
  gene_ids <- unique(gene_map$ENTREZID)
  universe_ids <- unique(universe_map$ENTREZID)

  # Skip groups that are too small after mapping.
  if (length(gene_ids) < min_genes_for_go || length(universe_ids) == 0) {
    term_counts[[group_name]] <<- c(before=0, after=0)
    cat("Skipping GO for", group_name, "- mapped genes:", length(gene_ids), "\n")
    return(data.frame())
  }

  # Test all eligible Biological Process terms using the tested-gene background.
  ego <- suppressMessages(clusterProfiler::enrichGO(
    gene = gene_ids,
    universe = universe_ids,
    OrgDb = org.Dm.eg.db,
    keyType = "ENTREZID",
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    readable = FALSE
  ))

  # Convert the enrichment result into a normal data frame.
  out <- as.data.frame(ego)
  terms_before <- nrow(out)
  if (terms_before == 0) { term_counts[[group_name]] <<- c(before=0, after=0); return(data.frame()) }
  ego_sig <- ego
  ego_sig@result <- ego@result %>% dplyr::filter(!is.na(p.adjust) & p.adjust <= go_fdr)
  if (nrow(ego_sig@result) == 0) {
    out_simplified <- data.frame()
    terms_after <- 0L
  } else {
    ego_sim <- suppressMessages(enrichplot::pairwise_termsim(ego_sig))
    ego_simplified <- suppressMessages(clusterProfiler::simplify(ego_sim, cutoff=0.7, by="p.adjust", select_fun=min, semData=go_semdata))
    out_simplified <- as.data.frame(ego_simplified)
    terms_after <- nrow(out_simplified)
  }
  term_counts[[group_name]] <<- c(before=terms_before, after=terms_after)
  full_out <- out
  full_out <- full_out %>% dplyr::mutate(Group=group_name, significant_BH=!is.na(p.adjust) & p.adjust <= go_fdr, geneSymbols=vapply(geneID, collapse_gene_symbols, character(1)))
  out <- out_simplified

  # Return a stable empty result when no ontology terms were testable.
  if (nrow(out) == 0) return(data.frame())

  # Add the biological group, significance flag, and readable gene symbols.
  out <- out %>%
    dplyr::mutate(
      Group = group_name,
      significant_BH = !is.na(p.adjust) & p.adjust <= go_fdr,
      geneSymbols = vapply(geneID, collapse_gene_symbols, character(1))
    ) %>%
    dplyr::arrange(dplyr::desc(Count), p.adjust, Description)

  # Export every tested term so non-significant results remain available for audit.
  safe_write_csv(
    full_out,
    file.path(go_output_dir, paste0("GO_BP_", group_name, "_ALL_tested_terms2.csv")),
    row.names = FALSE
  )

  # Export the simplified table used for summaries and figures.
  safe_write_csv(out, file.path(go_output_dir, paste0("GO_BP_", group_name, "_SIMPLIFIED_terms2.csv")), row.names=FALSE)

  # Export every significant term without applying an arbitrary top-N limit.
  safe_write_csv(
    dplyr::filter(out, significant_BH),
    file.path(go_output_dir, paste0("GO_BP_", group_name, "_ALL_significant_terms2.csv")),
    row.names = FALSE
  )

  # Return the complete table for summaries, detailed exports, and plotting.
  out
}

# Run GO enrichment for the three groups with one reusable call.
go_results <- lapply(
  names(group_maps),
  function(group_name) run_go(group_maps[[group_name]], group_name)
)

# Restore group names after lapply so later code is self-documenting.
names(go_results) <- names(group_maps)

# Combine every tested term from all non-empty group results.
all_go_terms <- dplyr::bind_rows(go_results)

# Keep every BH-significant term when enrichment returned results.
significant_go_terms <- if (nrow(all_go_terms) > 0) {
  dplyr::filter(all_go_terms, significant_BH)
} else {
  data.frame()
}

# Export one master table containing all tested terms from all three groups.
safe_write_csv(
  all_go_terms,
  file.path(go_output_dir, "GO_BP_three_groups_ALL_tested_terms2.csv"),
  row.names = FALSE
)

# Export one master table containing all significant terms from all three groups.
safe_write_csv(
  significant_go_terms,
  file.path(go_output_dir, "GO_BP_three_groups_ALL_significant_terms2.csv"),
  row.names = FALSE
)

# Build a concise group-level summary for interpretation and quality control.
go_summary <- dplyr::bind_rows(lapply(names(go_gene_sets), function(group_name) {
  data.frame(
    Group = group_name,
    input_genes = length(go_gene_sets[[group_name]]),
    mapped_genes = dplyr::n_distinct(group_maps[[group_name]]$gene),
    terms_before_simplification = unname(term_counts[[group_name]]["before"]),
    terms_after_simplification = unname(term_counts[[group_name]]["after"]),
    tested_GO_terms = nrow(go_results[[group_name]]),
    significant_GO_terms = if (nrow(go_results[[group_name]]) > 0) {
      sum(go_results[[group_name]]$significant_BH)
    } else {
      0
    }
  )
}))

# Display and export the GO analysis summary.
knitr::kable(go_summary, caption = "GO enrichment summary for all three groups")
safe_write_csv(
  go_summary,
  file.path(go_output_dir, "GO_BP_three_groups_summary2.csv"),
  row.names = FALSE
)
```

---

## List Every Significant GO Term

The table below includes every BH-significant term. The CSV exports also retain
all tested non-significant terms for complete review.

```{r all_go_terms_table}
# Select the most useful review columns while retaining every significant row.
go_review_table <- if (nrow(significant_go_terms) > 0) {
  significant_go_terms %>%
    dplyr::select(
      Group,
      ID,
      Description,
      GeneRatio,
      BgRatio,
      Count,
      pvalue,
      p.adjust,
      qvalue,
      geneSymbols
    )
} else {
  data.frame(Note = "No GO terms passed the configured BH threshold.")
}

# Use a paginated table in HTML so long results remain searchable and readable.
if (knitr::is_html_output()) {
  rmarkdown::paged_table(go_review_table, options = list(rows.print = 20))
} else {
  knitr::kable(go_review_table, digits = 4)
}
```

---

## Export Every GO-Term-to-Gene Link

```{r go_term_gene_details}
# Expand one significant GO result into one row per contributing original gene.
expand_go_genes <- function(go_table, mapped_genes, group_name) {
  # Return an empty table when this group has no significant terms.
  if (nrow(go_table) == 0 || !any(go_table$significant_BH)) return(data.frame())

  # Keep all significant terms and expand their slash-delimited Entrez lists.
  dplyr::bind_rows(lapply(
    which(go_table$significant_BH),
    function(i) {
      # Split the current GO term into individual Entrez identifiers.
      term_entrez <- strsplit(go_table$geneID[i], "/", fixed = TRUE)[[1]]

      # Link each Entrez ID back to every matching original gene identifier.
      mapped_genes %>%
        dplyr::filter(ENTREZID %in% term_entrez) %>%
        dplyr::distinct(gene, symbol_guess, ENTREZID) %>%
        dplyr::mutate(
          Group = group_name,
          GO_ID = go_table$ID[i],
          GO_Description = go_table$Description[i],
          GO_p_adjust = go_table$p.adjust[i],
          .before = 1
        )
    }
  ))
}

# Expand every significant term from all three groups with the same helper.
go_term_gene_details <- dplyr::bind_rows(lapply(
  names(go_results),
  function(group_name) expand_go_genes(
    go_results[[group_name]],
    group_maps[[group_name]],
    group_name
  )
))

# Sort the detailed table only when at least one term-to-gene link exists.
if (nrow(go_term_gene_details) > 0) {
  go_term_gene_details <- go_term_gene_details %>%
    dplyr::arrange(Group, GO_p_adjust, GO_ID, gene)
}

# Export the complete GO-term-to-gene audit table.
safe_write_csv(
  go_term_gene_details,
  file.path(go_output_dir, "GO_BP_ALL_significant_terms_with_ALL_genes2.csv"),
  row.names = FALSE
)

# Display the detailed table in a paginated format when rows are available.
if (nrow(go_term_gene_details) > 0 && knitr::is_html_output()) {
  rmarkdown::paged_table(go_term_gene_details, options = list(rows.print = 20))
} else if (nrow(go_term_gene_details) > 0) {
  knitr::kable(go_term_gene_details, digits = 4)
} else {
  cat("No significant GO-term-to-gene links were available.\n")
}
```

---

## Visualise Every Significant GO Term

Each bar represents one significant GO term. Bar height shows `Count`, the
number of genes from the analysed group associated with that term. Terms run
from largest to smallest from left to right. The plots are divided into pages
of 15 terms so the 14-point x-axis labels remain readable.

```{r go_plots, fig.width=18, fig.height=10}
# Create paginated bar charts ordered by descending GO-term gene count.
plot_go_bars <- function(go_table, group_label, terms_per_page = 15) {
  # Stop cleanly when enrichment did not return a result table.
  if (is.null(go_table) || nrow(go_table) == 0) {
    cat("No GO results for", group_label, "\n")
    return(invisible(NULL))
  }

  # Retain every significant term and rank terms from largest to smallest.
  significant_terms <- go_table %>%
    dplyr::filter(significant_BH) %>%
    dplyr::arrange(dplyr::desc(Count), p.adjust, Description)

  # Stop cleanly when no term passes the selected BH threshold.
  if (nrow(significant_terms) == 0) {
    cat("No significant GO terms for", group_label, "\n")
    return(invisible(NULL))
  }

  # Store the descending size rank used to paginate the complete term list.
  significant_terms$Size_rank <- seq_len(nrow(significant_terms))

  # Divide the ordered terms into readable pages without dropping any term.
  page_ids <- ceiling(significant_terms$Size_rank / terms_per_page)

  # Build one bar chart for each consecutive size-ranked page.
  for (page_id in unique(page_ids)) {
    # Select the next group of largest remaining GO terms.
    page_terms <- significant_terms[page_ids == page_id, , drop = FALSE]

    # Wrap long term names while preserving the descending left-to-right order.
    page_terms$Term <- paste0(
      page_terms$ID,
      "\n",
      stringr::str_wrap(page_terms$Description, width = 24)
    )
    page_terms$Term <- factor(page_terms$Term, levels = page_terms$Term)

    # Convert adjusted p-values to a clear positive colour scale.
    page_terms$MinusLog10FDR <- -log10(
      pmax(page_terms$p.adjust, .Machine$double.xmin)
    )

    # Draw the bar chart with GO terms on the x-axis and gene count on the y-axis.
    bar_plot <- ggplot2::ggplot(
      page_terms,
      ggplot2::aes(x = Term, y = Count, fill = MinusLog10FDR)
    ) +
      ggplot2::geom_col(width = 0.78) +
      ggplot2::geom_text(
        ggplot2::aes(label = Count),
        vjust = -0.35,
        size = 4.5
      ) +
      ggplot2::scale_fill_gradient(low = "#BFD3E6", high = "#084594") +
      ggplot2::scale_y_continuous(
        expand = ggplot2::expansion(mult = c(0, 0.12))
      ) +
      ggplot2::theme_minimal(base_size = 14) +
      ggplot2::theme(
        panel.grid.major.x = ggplot2::element_blank(),
        panel.grid.minor = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_text(
          size = 14,
          angle = 60,
          hjust = 1,
          vjust = 1
        ),
        axis.text.y = ggplot2::element_text(size = 14),
        axis.title = ggplot2::element_text(size = 16),
        plot.title = ggplot2::element_text(size = 18, face = "bold"),
        plot.subtitle = ggplot2::element_text(size = 14),
        legend.title = ggplot2::element_text(size = 14),
        legend.text = ggplot2::element_text(size = 13)
      ) +
      ggplot2::labs(
        title = paste("GO Biological Process enrichment:", group_label),
        subtitle = paste0(
          "Page ", page_id, " of ", max(page_ids),
          "; size ranks ", min(page_terms$Size_rank),
          "\u2013", max(page_terms$Size_rank),
          " (largest to smallest by gene count)"
        ),
        x = "GO term",
        y = "Gene count",
        fill = expression(-log[10]("BH adjusted p-value"))
      )

    # Print the current bar-chart page into the knitted report.
    print(bar_plot)
  }
}

# Plot all significant majority-supported DE GO terms as size-ranked bar charts.
plot_go_bars(go_results$majority_supported_DE, "majority-supported DE genes")

# Plot all terms from genes repeatedly higher in D. erecta.
plot_go_bars(
  go_results$upregulated_in_Dere,
  "upregulated in D. erecta"
)

# Plot all terms from genes repeatedly lower in D. erecta.
plot_go_bars(
  go_results$downregulated_in_Dere,
  "downregulated in D. erecta"
)
```


## Reproducibility
<details><summary>Session information</summary>

```{r session_info}
sessionInfo()
```

</details>
